Gap formation and stability in non-isothermal protoplanetary discs

### 2014-11-21 paper/results.tex
major updates by MKL

\section{Growth of non-axisymmetric modes without the influence of the
  planet}\label{linear1}
In this section, the planet is introduced at $t=20P_0$ and 
its potential switched on over $10P_0$. At $t=30P_0$ we switch off the
planet potential and azimuthally average the surface density, energy
and velocity fields. At this point the planet has carved a partial
gap and the RWI has not yet occured.
 We then perturb the surface density for $r>r_p$ and continue to 
evolve the disc. We impose  sinusoidal perturbations with 
azimuthal wavenumbers $m\in[1,10]$ and random amplitudes within $\pm 0.01$
 times the local surface density.% {\bf what are PERTMIN and PERTMAX values?}  
This procedure allows us to analyse the growth of 
non-axisymmetric modes associated with the gap, but without
complications from non-axisymmetry arising directly from disc-planet
interaction (i.e. planet-induced wakes).

@@

Simulations here employ a resolution of $(N_r,N_{\phi})=(1024,2048)$
with open boundaries at $r=r_\mathrm{in}$ and
$r_\mathrm{out}=25r_\mathrm{in}$. We consider   
%for a time up to $200P_0$ allowing high resolution and detection of high
%$m$ structures.
cases of $\tilde{\beta}=0.1,1,10$ corresponding to fast, moderate,
and slowly cooled discs. % {\bf boundary condition for these sims?}

@@

%For all cases the gap width is $\approx 5r_{hill}$
%which is set by torque balances in the disk \citep{crida06}.

Larger $\tilde\beta$ values result in higher disc aspect ratios $h=H/r$,
i.e. higher temperatures. Heating mostly occur at the gap edges
due to planet-induced spiral shocks. Decreasing the cooling time
implies that this heat is retained in the disc. In the inviscid limit the gap
opening condition is $r_h\gtrsim H$ or $q\gtrsim 3h^3$
\citep{crida06}. 
% which is a value directly correlated to temperature in our model. 
%For a planet to carve a gap in the surrounding material it has been
%shown that $q>H^3$ is needed \citep{crida06}. 
This indicates that for hotter discs (higher
$h$), it becomes more difficult for a planet of fixed $q$ to open a
gap. This explains the shallower gaps in surface density when
$\tilde{\beta}$ is increased. 
%However, notice the gap structure
%changes most significantly for $\beta =0.1 \to 1.0$. For $\beta \geq
%1$

%This is seen in our simulations as changes in $\tilde{\beta}$
%result in changes of $h$ by less than $20\%$ yet gap depth almost
%doubles.

@@

%Charactereristicly
%for such disk-planet Rossby vortices the generalized vortensity
%minima correspond to density maxima in the disk. 
%Since generalized vortensity extrema are also
% correspondingly less pronounced and density gradients become smaller
% we expect that the stability of gaps to increase with larger
% $\tilde{\beta}$.

@@

\subsection{Axisymmetric stability}
%Check that the gap profiles are stable against axisymmetric
%stability. Can use analytical criteria given in section 3.1 of Li et
%al (2000, ApJ, 533, 1023). No plot needed, just statements. 
%}
The intial planet-disc interaction form bumps and grooves in the gap profiles
which can potentially be unstable due to axisymmetric instabilities. The
generalised local axisymmetric stability condition is the Solberg-Hoiland
criterion,  
\begin{align}
  \kappa^2+N^2 \geq 0 
\end{align}
where
\begin{align}
 N^2=\frac{1}{\Sigma} \frac{\partial P}{\partial r}
 \left(\frac{1}{\Sigma} \frac{\partial \Sigma}{\partial
     r}-\frac{1}{\gamma P} \frac{\partial P}{\partial r}  \right) 
\end{align}
is the square of the Brunt-V\"ais\"al\"a frequency.  
%for the disk due to entropy
%variation which is non-zero for non-adiabatic models such as being
%considered. 
At the outer gap edge $r=r_p+2.5r_h$,  where the RWI is excited
(see below), we find $\kappa^2 + N^2$ reaches local minimum with a value
$\sim 0.002$ (code units) for all $\tilde\beta$. The
Brunt-V\"ais\"al\"a frequency is $N\sim 10^{-5}$. The Solberg-Hoiland
criteria is similary satisfied for the entire disc throughout the
simulations. % with little change of values. 
Thus for all values of $\tilde\beta$ the planet-induced gaps are
stable to axisymmertic instabilities.

\subsection{Non-axisymmetric instability}\label{linear}
%{\bf present `planet-off' simulations. one `fourier mode v.s. time'
%  plot to show growth of linear instability. do an adiabatic case (or
%  extremely long cooling time) to see the effect of heated gap edge
%  (i.e. temp doesn't go back to t=0 value too quickly)
%  compare growth rate and dominant
%  m as function of beta (table). 2D figs to contrast (also used to
%  show it's the minimum in generalized pv that goes unstable.    
%  result: increasing cooling time makes the gap
%  more stable, and favors lower m. note: the `basic state' should be
%  the system at t=30 after azimuthal average. linear results should
%  have small perturbations. 
%}

%After the azimuthal averaging, the small pertibation excites rapidly
%growing instabilities by the RWI.

We now examine the evolution of the gap for $t>30P_0$, with the
planet potential switched off, but with an added surface density
perturbation. For all three cooling times $\tilde{\beta}=0.1,\, 1.0,\,
10$, we observe exponential growth of non-axisymmetric
structures. % after the 
%planet potential is switched off and the system subject to random
%perturbations.  
An example is shown in Fig. \ref{linearmodes} for 
$\tilde{\beta}=10$. We charcterize these
modes with an azimuthal wavenumber $m$ and growth rate $\gamma(m)$ as defined by
Eq.~\ref{fouriertransform}---\ref{growth}. Mode amplitudes were
averaged over $r-r_p\in[2,5]r_h$. Table \ref{modetable}
lists the growth rates measured during 
linear growth for 5 values of $m$ centred around the maximum.

@@

% which result in hotter gaps since the heating
%due
%This is as expected from the
%intial gap profiles, because 
% and the gap forming criteria discussed in the
%previous section.

\begin{figure}
  \centering
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity50}
  }
\hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity67}
  }
\hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity75}
  }  
  \caption{Generalized vortensity perturbation (relative to $t=0$) for
    cases of $\tilde{\beta}=0.1,1,10$ (left,middle,right) during
    the growth of non-axisymmetric modes. The planet potential has
    been switched off.  The number of vortices
    decrease as $\tilde{\beta}$ increases. Note that snapshots are
    taken later for increasing $\tilde{\beta}$ because it takes longer
    for the vortices to grow and become visible with increasing cooling time. 
    % correspondingly increase to capture the longer growth phases of
    % the instabilities. 
    \label{2Dlinear} 
%{\bf I'm curious to know if plots clearer if
%      perturbations are measured relative to t=30? remember t=30 is
%      our reference state for instability. But we'll probably keep
%      these for consistency (later non-linear plots are all measure
%      with respect to t=0).}
  } 
\end{figure}

\subsubsection{Nearly adiabatic discs}
\label{adiabatic_section}
%{\bf a case with extrmely long cooling time (or adiabatic run) in
%  order to look at effect of heated gap edge.  prelim result: not
%  important, at t=30, gap edge only heats to $H/r\simeq0.06$ even for
%  purely adiabatic disk. change in $h$ not important for linear
%  perturbations (but important for setting up the basic state). 
%}
The above `planet-off' simulations are not formally linear
stability calculations, because the cooling time is shorter
than the instability growth time, 
%
%as we do not capture the effects of a heated
%disk on the instability growth. 
$t_c<\gamma^{-1}$.  
Thus the disk cools back to its intial temperature corresponding to
$h=0.05$ before or during the instability growth, so we do not
have a steady basic state to formulate a standard linear stability 
problem. %This means the disc temperture of $h=0.05$ in the the above cases

In order to perform a proper linear stability analysis and capture the
effect of a heated gap edge during instability growth, we ran a simulation  with
$\tilde{\beta}=100$, corresponding to an almost adiabtaic disc.  
%To compare our 'planet-off' analysis with a true
%linear stability analysis 
In this simulation the cooling rate is slow enough that the gap 
temperature profile (e.g. middle panel of Fig. \ref{intial1D}) changes
only marginally over the instability growth timescale. %heat is
                                %retained/maintained.

We find very similar gap profiles and mode growth rates for
$\tilde{\beta}=100$ as with $\tilde{\beta}=10$. The disc heats up to
values $h\simeq0.06$ in the nearly adiabatic case. This is close to
the original temperature of $h=0.05$, so linear growth rates are not expected
to change significantly %according to the linear theory presented in
\citep{li00}.

@@

%results in shallower gaps
%which is expected to be

%the difference in growth rates between this and the
%original disc temperature of $h=0.05$ changes almost insignificantly
%as shown by linear theory \citep{li00}. 
%Although linear
%theory predicts higher growth rates with increasing $h$

%This indicate that a
%continuously hot gap edge is not as important for development of
%vortices as the heating effects on the formation of the intial gap
%state.

%{\bf
%If there's time, do low resolution planet off simulations but keep the planet on
%until $t=40P_0$ or $50P_0$. Look at stability properties when the gaps
%are generally deeper (including the nearly adiabatic case). Although
%with the planet instability may already appear by $t=40P_0$, the
%azimtuhal average performed prior to perturbations will erase it.}

@@

%Similar
%quasi-steady vortices form as in the cases with continuous planet interaction
%but the long-term behaviour is significantly
%different without the planet as seen in 
%The quasi-steady vortices formed have significantly lower density pertibations
%and now slowly decrease in amplitute in time as apposed to previously
%where their growth was supported by planet interactions. 
%The vortices simply
%dissipate viscously with dissipation rate decreasing with larger $\tilde\beta$.

%Non-linear observations were also made for the `planet-off'
%simulations.
% as the vortices grow as linear
% instabilities discussed previously.

%The vortices are never found to undergo the drastic sudden dissipation effects
%that characterize the planet induced growing cases do.

@@

\section{Non-linear evolution of
  gap-edge vortices with finite cooling time}\label{nonlinear} 
%{\bf main fig: vortex amplitude v.s. time for diff beta. enough data
%  to plot vortex lifetime v.s. beta? table: 
%  averaged quantities over quasi-steady state: aspect-ratio (to
%  compare with fu at al), rossby number (vortex strength
%  v.s. cooling?), maybe alpha visc. vortex size: visible difference?  
%  only inviscid cases. describe evolution of one case. main
%  conclusion: longer vortex lifetime with increasing cooling time (up
%  to some optimal timescale). vortex death: induced-shock and/or
%  smoothing the gap edge. describe simulation setup, resolution?
%  should mention that results consistent with lower-resolution prelim
%  runs. maybe torques? 
%}

We now examine long-term simulations of gap-edge vortices for
$\tilde{\beta}=0.1,0.5,1,5,10$. The planet potential is kept on
throughout.  
% up to a total of
%$2000P_0$. 
%Planets were left free to interact dynamiclly with the disk
%and vortices after $t=30P_0$ as apposed to previous section.  
We employ a grid with $(N_r,N_{\phi})=(512,1024)$ in order for these
simulations to be computationally feasible. We also use a larger
disc with $r_{\mathrm{out}}=45r_{\mathrm{in}}$ to minimize boundary
effects on vortex evolution, and apply open boundaries at
$r=r_\mathrm{in},\,r_\mathrm{out}$. We comment that lower-resolution
simulations with $(N_r,N_{\phi})=(256,512)$ were initially carried
out, which show similar behaviour and trends as the high-resolution
runs reported below.

%where also done for the above $\tilde\beta$ values
%with the additional inclusions of $\tilde{\beta}=0.25,0.75,2.5,7.5$.
%Vortex evolution and behaviour are convergent with
%higher resolution simulations as discussed in this section.
% {\bf mention lower resolution preliminary runs - 
%similar results to those presented here (?). want to demonstrate convergence}

%An open boundry
%condition was used in the outer edge of
% allowing significant room for
%linblad resonances of the vortices.

\subsection{Generic evolution} 
The linear growth of the RWI and vortex-formation is followed by 
vortex merging. We now find merging timescales independent of
$\tilde\beta$, and by $60P_o$ only one vortex remains.  
%as the growth of these modes are now 
%accelerated by planetary influence as apposed to `planet off' simulations.
% i.e. only the $m=1$ mode remains. 
The evolution of the amplitude of the $m=1$ surface density component,
averaged over $r-r_p\in[2,10]r_h$ is shown in Fig~\ref{lifetimeplot} 
for different $\tilde\beta$. 
{\bf any particular reason for the larger radial range for average?
  did we get bigger vortices?} 
% {\bf should state (also for planet-off runs)
% what the radial range is used for averaging the Fourier amplitudes}

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
  0cm 0cm 1cm]{figures/longterm_stability}
  \caption{Evolution of the $m=1$ surface density component,
    non-dimenionlised by the initial $m=0$ component, for long term
    simulations with the planet potential kept on. The case with 
    $\tilde\beta=5.0$ (red) gave the longest-living vortex. {\bf
      remember to update to 
      include vortex dissipation for all cases} \label{lifetimeplot}} 
\end{figure}

In all cases the system remains in a quasi-steady state for
$\gtrsim800P_0$ with a single vortex circulating 
the outer gap edge at the local Keplerian  
frequency. Fig. \ref{Vortex2D} shows a typical plot of the relative 
surface density perturbation in this state. During this stage, the 
vortex intensifies. As the vortex forms it has an initial 
Rossby number $Ro\approx-0.1$ for all $\tilde\beta$
(indicating anti-cyclonic
motion). As the $m=1$ amplitude grows the Rossby number also increases
in magnitude, 
%relative spin of the vortex
%continuously grows along with it
, reaching a characteristic value of $Ro\approx-0.4$ for all cooling
rates. {\bf is the the value reached just before decay for all cooling?}

The surface density perturbation (relative to $t=0$)  
can reach values of $\Delta\Sigma/\Sigma_0 \sim 7$ at the center of vortices
for all cases of $\tilde\beta$ in quasi-steady state, and
was found as high as $\Delta\Sigma/\Sigma_0 \sim 9$ for the 
$\tilde\beta=10$ case. This large increase in surface density is due 
vortex growth, since there is continuous generation of vorticity by
planet-disc interaction. This is supported by the observation that in
the previous simulations without the planet, the amplitude of
the post-merger vortex does not grow (Fig
\ref{planetofflifetimeplot}).  We remark that the increase in the
vortex surface density directly due to the pile-up of material at the
gap edge (because of gap-opening) is small: the average surface density
perturbation near the outer gap edge, $r\sim r_{p}+2.5r_h$, is 
$\langle\Delta\Sigma/\Sigma_0\rangle_\phi<1$.

%{\bf is vortex merging affected by $\tilde{\beta}$? e.g. timescale
%  taken to reach one vortex as a function of cooling}
%{\bf for what cooling are these values quoted for? do
%  these values depend on cooling?}
% {\bf is the evidence for this the fact that growth is not observed in
%  planet-off simulations?}   
%{\bf note that there
%  are two origins for the surface density boost relative to $t=0$: 1) gap
%  formation, which increases the surface density at the location of
%  the instability (since they happen at surface density max., and 2)
%  vortex instability). what is the surface density boost relative to
%  the gap profile?}

Interestingly, in most cases we observe a sudden drop in 
the $m=1$ amplitude within the simulation time, which occurs over
timescale between $20P_0$ to $70P_0$, i.e. much shorter than the duration of the
quasi-steady state. We did not find this dissipation timescale to
correlate with $\tilde\beta$. %but instead are paramtrized by vortex Mach numbers.
%{\bf is this dissipation timescale independent of cooling? Trend is
%based on mach values which dont follow linear trend }  
%Eventually
%the $m=1$ amplitudes are also characterized by 
%After these characteristic drops in vortex amplitude, the instability
%is never seen to reform at such large overdensities: the vortex surface
%densities are never larger than that due to the planetary wakes. 
We designate the time at which these amplitude drops as the lifetime 
of the quasi-steady vortex. After the vortices decay the RWI is 
not seen to be excited again and no vortices are seen to refrom within the
simulations time. 
{\bf Perhaps a snapshot of what the gap edge looks like after vortex death.
}
%The instability is not seen to emerge again after the intial vortex dissipation
% due to the smoothening out of the gap edge during dissipation by the
% transfer of density from the once tightly held vortex region to the surrounding disk and gap.  
% {\bf Discussed in next section:  question: does instability
%   actually happen again? if it does, but just weaker, that may be
%   because the first vortex has smoothed out the gap edge, so
%   instabilities following that would be weaker.}

@@

\subsection{Additional analysis on vortex dissipation}
{\bf are these high resolution results? vortex lifetime in snapshots
  not consistent with amplitude plots}
%Conjecture: vortex death due to vortex shocking the system and
%dissipating its energy. discuss one case. here we want to explain or
%suggest why vortices die} 
%Rossby values have dynamics that are coupled with the lifetime and
%$m=1$ amplitude of the vortices.
In this subsection we examine the sudden vortex decay observed in our
simulations in more detail. Fig. \ref{shockplot} show snapshots
of the vortex leading up to its decay for the case $\tilde{\beta}=1$. 
The surface density perturbation and the surface density gradient are
shown. In quasi-steady ($t=700P_0$) the vortex is elongated, but becomes
more compact as it grows before decaying. {\bf rough estimate of
  vortex aspect-ratio in the three snapshots? 
  (length/width, by inspection is ok)}

Notice in Fig. \ref{shockplot} the appearence of wakes extending from
either side of the vortex at $t=1300P_0$ as it grows. These 
wakes correlate with large gradients in surface density (bottom
panel), and are first seen in the later half of the quasi-steady state.       
% As the vortices grow in intensity wakes are seen to extend from them. 
During this phase the average value of the surface density gradient
along the wakes is $|\nabla\Sigma/\Sigma| \sim 4 $.{\bf quote something
  dimensionless, or say `in code units'}  
Just before the vortices start to decay, we observe this quantity
sharply increases to $ \sim 6 $, and remains around this value
until the vortex dies out, at which point the 
associated Rossby number quickly approaches zero.

\begin{figure}
\subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock1}
  }
\hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock2}
  }
\hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock3}
  } \\[-0.98cm]
\subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock4}
  }
\hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock5}
  }
\hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock6}
  } 
  \caption{Evolution of the vortex in the case $\tilde{\beta}=1$
    just before its sudden decay. The surface density perturbation
    (top) and the associated surface density gradient (bottom) are shown.
    \label{shockplot}}
  % Large surface density gradients become prominant
  % around vortex during dissipation} 
\end{figure}

We also measured large increases in the Mach number near
the vortex when it begins to decay. Figure \ref{machplot} plots the
Mach number $M=|\bm{v} - \bm{v}_\mathrm{vor}|/c_s$, where
$\bm{v}_\mathrm{vor}$ corresponds to the bulk velocity of the vortex
around the disc. {\bf check - i assume this is the definition of mach
  number used here}. Values in Fig. \ref{machplot} have been averaged over
a region within $2H$ of the vortex centre.    
% Mach numbers with repect to the vortex bulk motion around the vortex
% region of $|r-r_{vortex}|<2H$
% are shown in  for various cooling rates over time.
During the quasi-steady state the Mach number increase
steadily, and for all cases $M$ reaches a maximum at 
% with larger growth rate for higher $\tilde\beta$.
a time coincident with the start of vortex decay.

Putting the above observations together, we suggest the sudden decay
in the vortex amplitude is due to shock formation \emph{by the
  vortex}. When the vortex reaches large amplitude, it begins
to induce shocks in the surrounding fluid, as supported by the
increase in Mach number and the appearence of wakes with large surface
density gradients. The vortex may then lose energy through shock
dissipation. In addition, a strong vortex (or shock formation) 
also smoothes out the gap structure that originally gave rise to the
RWI, this would oppose vortex growth. We examine this below.

\begin{figure}
    \includegraphics[width=\linewidth]{figures/mach}
 \caption{Running-time average {\bf ?} of the Mach number relative to
   the vortex, averaged over a region within $2H$ of the  
   vortex centre. This plot can be compared to the evolution of the
   vortex amplitude shown in Fig. \ref{lifetimeplot}.
   {\bf remember to update plot to include full simulation duration}
   \label{machplot}}
\end{figure}

>>>>

\section{Growth of non-axisymmetric modes without the influence of the
  planet}\label{linear1}
In this section, the planet is introduced at $t=20P_0$ and 
its potential switched on over $10P_0$. At $t=30P_0$ we switch off the
planet potential and azimuthally average the surface density, energy
and velocity fields. At this point the planet has carved a partial
gap and the RWI has not yet occured.
 We then perturb the surface density for $r>r_p$ and continue to 
evolve the disc. We impose  sinusoidal perturbations with 
azimuthal wavenumbers $m\in[1,10]$ and random amplitudes within $\pm 0.01$
 times the local surface density. 
This procedure allows us to analyse the growth of 
non-axisymmetric modes associated with the gap, but without
complications from non-axisymmetry arising directly from disc-planet
interaction (i.e. planet-induced wakes).

@@

Simulations here employ a resolution of $(N_r,N_{\phi})=(1024,2048)$
with open boundaries at $r=r_\mathrm{in}$ and
$r_\mathrm{out}=25r_\mathrm{in}$. We consider   
cases with $\tilde{\beta}=0.1,1,10$ corresponding to fast, moderate,
and slowly cooled discs. % {\bf boundary condition for these sims?}

@@

Larger $\tilde\beta$ values result in higher disc aspect ratios $h=H/r$,
i.e. higher temperatures. Heating mostly occur at the gap edges
due to planet-induced spiral shocks. Decreasing the cooling time
implies that this heat is retained in the disc. In the inviscid limit the gap
opening condition is $r_h\gtrsim H$ or $q\gtrsim 3h^3$
\citep{crida06}. 
This indicates that for hotter discs (higher
$h$), it becomes more difficult for a planet of fixed $q$ to open a
gap. This explains the shallower gaps in surface density when
$\tilde{\beta}$ is increased. 
%However, notice the gap structure
%changes most significantly for $\beta =0.1 \to 1.0$. For $\beta \geq
%1$

@@

@@

\subsection{Axisymmetric stability}
The intial planet-disc interaction form bumps and grooves in the gap profiles
which can potentially be unstable due to axisymmetric instabilities. The
generalised local axisymmetric stability condition is the Solberg-Hoiland
criterion,  
\begin{align}
  \kappa^2+N^2 \geq 0 
\end{align}
where
\begin{align}
 N^2=\frac{1}{\Sigma} \frac{\partial P}{\partial r}
 \left(\frac{1}{\Sigma} \frac{\partial \Sigma}{\partial
     r}-\frac{1}{\gamma P} \frac{\partial P}{\partial r}  \right) 
\end{align}
is the square of the Brunt-V\"ais\"al\"a frequency. 
At the outer gap edge $r=r_p+2.5r_h$,  where the RWI is excited
(see below), we find $\kappa^2 + N^2$ reaches local minimum with a value
$\sim 0.002$ (code units) for all $\tilde\beta$. The
Brunt-V\"ais\"al\"a frequency is $N\sim 10^{-5}$. 
{\bf Brunt frequency the same for all cooling?}
The Solberg-Hoiland criteria is similary satisfied for the entire disc throughout the 
simulations. {\bf i assume this means the azimuthally averaged
  values. do you ever get $N^2<0$ anywhere in the 2D disc?}  
Thus for all values of $\tilde\beta$ the planet-induced gaps are
stable to axisymmertic instabilities.

\subsection{Non-axisymmetric instability}\label{linear}
We now examine the evolution of the gap for $t>30P_0$, with the
planet potential switched off, but with an added surface density
perturbation. For all three cooling times $\tilde{\beta}=0.1,\, 1.0,\,
10$, we observe exponential growth of non-axisymmetric
structures.  
An example is shown in Fig. \ref{linearmodes} for 
$\tilde{\beta}=10$. We charcterize these
modes with an azimuthal wavenumber $m$ and growth rate $\gamma(m)$ as defined by
Eq.~\ref{fouriertransform}---\ref{growth}. Mode amplitudes were
averaged over $r-r_p\in[2,5]r_h$. Table \ref{modetable}
lists the growth rates measured during 
linear growth for 5 values of $m$ centred around the maximum.

@@

\begin{figure}
  \centering
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity50}
  }
\hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity67}
  }
\hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity75}
  }  
  \caption{Generalized vortensity perturbation (relative to $t=0$) for
    cases of $\tilde{\beta}=0.1,1,10$ (left,middle,right) during
    the growth of non-axisymmetric modes. The planet potential has
    been switched off.  The number of vortices
    decrease as $\tilde{\beta}$ increases. Note that snapshots are
    taken later for increasing $\tilde{\beta}$ because it takes longer
    for the vortices to grow and become visible with increasing cooling time. 
    \label{2Dlinear} 
  } 
\end{figure}

\subsubsection{Nearly adiabatic discs}
\label{adiabatic_section}

The above `planet-off' simulations are not formally linear
stability calculations, because the cooling time is shorter
than the instability growth time, 
$t_c<\gamma^{-1}$.  
Thus the disk cools back to its intial temperature corresponding to
$h=0.05$ before or during the instability growth, so we do not
have a steady basic state to formulate a standard linear stability 
problem.

In order to perform a proper linear stability analysis and capture the
effect of a heated gap edge during instability growth, we ran a simulation  with
$\tilde{\beta}=100$, corresponding to an almost adiabtaic disc.  
In this simulation the cooling rate is slow enough that the gap 
temperature profile (e.g. middle panel of Fig. \ref{intial1D}) changes
only marginally over the instability growth timescale.

We find very similar gap profiles and mode growth rates for
$\tilde{\beta}=100$ as with $\tilde{\beta}=10$. The disc heats up to
values $h\simeq0.06$ in the nearly adiabatic case. This is close to
the original temperature of $h=0.05$, so linear growth rates are not expected
to change significantly \citep{li00}.

@@

@@

@@

\section{Non-linear evolution of
  gap-edge vortices with finite cooling time}\label{nonlinear} 
We now examine long-term simulations of gap-edge vortices for
$\tilde{\beta}=0.1,0.5,1,5,10$. The planet potential is kept on
throughout. We employ a grid with $(N_r,N_{\phi})=(512,1024)$ in order
for these 
simulations to be computationally feasible. We also use a larger
disc with $r_{\mathrm{out}}=45r_{\mathrm{in}}$ to minimize boundary
effects on vortex evolution, and apply open boundaries at
$r=r_\mathrm{in},\,r_\mathrm{out}$. We comment that lower-resolution
simulations with $(N_r,N_{\phi})=(256,512)$ were initially carried
out, which show similar behaviour and trends as the high-resolution
runs reported below.

\subsection{Generic evolution} 
The linear growth of the RWI and vortex-formation is followed by 
vortex merging. We now find merging timescales independent of
$\tilde\beta$, and by $60P_o$ only one vortex remains. 
The evolution of the amplitude of the $m=1$ surface density component,
averaged over $r-r_p\in[2,10]r_h$, is shown in Fig.~\ref{lifetimeplot} 
for different $\tilde\beta$. 
%{\bf any particular reason for the larger radial range for average?
%  did we get bigger vortices?:Yes vortices get quite large in quais steady,
% changing the planet off averaging range to [2,10] changes the plots negligibly
% if we want to just use this value range for consistancy }

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
  0cm 0cm 1cm]{figures/longterm_stability}
  \caption{Evolution of the $m=1$ surface density component,
    non-dimenionlised by the initial $m=0$ component, for long term
    simulations with the planet potential kept on.\label{lifetimeplot}}   
\end{figure}

In all cases the system remains in a quasi-steady state for
$\gtrsim800P_0$ with a single vortex circulating 
the outer gap edge at the local Keplerian  
frequency. Fig. \ref{Vortex2D} shows a typical plot of the relative 
surface density perturbation in this state. During this stage, the 
vortex intensifies. As the vortex forms it has an initial 
Rossby number $Ro\approx-0.1$ for all $\tilde\beta$
(indicating anti-cyclonic 
motion). As the $m=1$ amplitude grows the Rossby number also increases
in magnitude, 
%relative spin of the vortex
%continuously grows along with it
 reaching a characteristic minimum value of $Ro\approx-0.35$ to $-0.45$ for
all cooling rates. 
{\bf Maybe useful to have a Rossby number v.s. time plot to complement
  mode amplitude plot. 
}
%{\bf is the the value reached just before decay for all cooling?:yes 
%values peak in small range [-0.35,-0.45] for all beta}

We find the vortices can become significantly over-dense. 
The surface density perturbation (relative to $t=0$)  
can reach values of $\Delta\Sigma/\Sigma_0 \sim 7$ at the center of vortices
for all cases of $\tilde\beta$ in quasi-steady state, and
was found as high as $\Delta\Sigma/\Sigma_0 \sim 9$ for the 
$\tilde\beta=10$ case. {\bf any trend between maximum over-density
  achieved and cooling?}
This large increase in surface density is due to  
vortex growth, since there is continuous generation of vorticity by
planet-disc interaction. This is supported by the observation that in
the previous simulations without the planet, the amplitude of
the post-merger vortex does not grow (Fig
\ref{planetofflifetimeplot}).  We remark that the increase in the
vortex surface density directly due to the pile-up of material at the
gap edge (because of gap-opening) is small: the average surface density
perturbation near the outer gap edge, $r\sim r_{p}+2.5r_h$, is 
$\langle\Delta\Sigma/\Sigma_0\rangle_\phi<1$.

Fig. \ref{lifetimeplot} shows the duration of the quasi-steady state
varies with the cooling rate: for
$\tilde{\beta}=0.1$ and $10$, the vortex amplitude begins to decay around
$t\sim900P_0$, while for $\tilde{\beta}=1,\,5$ the decay begins at 
$t\sim1200P_0$. This non-monotonic dependence suggests that there
exists an optimal cooling rate to maximise the vortex lifetime. We
will discuss this issue futher in
\S\ref{lifetime_discuss}. However, the decay timescale decreases with
increasing cooling time: for $\tilde{\beta}=0.1$ it takes $\sim400P_0$
whereas for $\tilde{\beta}=10$ it takes $\sim 100P_0$ for the $m=1$
amplitude to decay significantly after reaching maximum.

%These
%timescales are typically much shorter than the duration of the
%quasi-steady state.

{\bf question: vortex amplitude plot suggest just after linear growth,
  vortex amplitudes are smaller with increasing beta, but later on in
  the quasi-steady state, this trend reverses?  
}
%The
%decay timescale appears to decrease with increasing cooling time. For
%$\tilde{}$  
%Interestingly, in most cases we observe a sudden drop in 
%the $m=1$ amplitude within the simulation time, which occurs over
%timescale between $20P_0$ to $70P_0$, i.e. much shorter than the duration of the
%quasi-steady state. 
%We did not find this dissipation timescale to
%correlate with $\tilde\beta$. 
%We designate the time at which these amplitude drops as the lifetime 
%of the quasi-steady vortex. After the vortices decay the RWI is 
%not seen to be excited again and no vortices are seen to refrom within the
%simulations time. 
{\bf Perhaps a 2D snapshot of what the gap edge looks like after
  vortex death (like the one in the poster). defer definition of
  vortex lifetime to later, when we discuss it. 
}

@@

\subsection{Additional analysis on vortex decay}
In this subsection we examine the vortex decay observed in our
simulations in more detail. Fig. \ref{shockplot} show snapshots
of the vortex for the case $\tilde{\beta}=1$. The plots show the surface
density perturbation and the surface density gradient during
quasi-steady state ($t=700P_0$), when the $m=1$ amplitude begins to
decrease ($t=1300P_0$) and just after the rapid amplitude decay
($t=1510P_0$).

In quasi-steady ($t=700P_0$) the vortex is elongated, but becomes
more compact as it grows before decaying. 
{\bf rough estimate of
  vortex aspect-ratio in the three snapshots? 
  (length/width, by inspection is ok)}  
Notice in Fig. \ref{shockplot} the appearence of wakes extending from
either side of the vortex at $t=1300P_0$. These 
wakes correlate with large gradients in surface density (bottom
panel), and are first seen in the later half of the quasi-steady
state. We find the time at which 
the vortex begins to decay coincides with the emergence of these
wakes. {\bf true statement?}

% As the vortices grow in intensity wakes are seen to extend from them. 
During quasi-steady state the average value of the surface density gradient
along the wakes is $|\nabla\Sigma/\Sigma| \sim 4 $.{\bf quote something
  dimensionless, or say `in code units'}  
Just before the $m=1$ amplitude begins to decrease, we observe this quantity
sharply increases to $ \sim 6 $, and remains around this value
until the vortex dies out, at which point the associated Rossby number
quickly approaches zero.

\begin{figure}
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock1}
  }
  \hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock2}
  }
  \hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock3}
  } \\[-0.98cm]
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock4}
  }
  \hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock5}
  }
  \hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock6}
  } 
  \caption{The vortex in the case $\tilde{\beta}=1$
    during quasi-steady state (left), at maximum amplitude of the
    $m=1$ surface density component, 
    (middle), and just after the decay in the $m=1$ amplitude
    (right). The surface density perturbation
    (top) and the associated surface density gradient (bottom) are shown.
    \label{shockplot}}
  % Large surface density gradients become prominant
  % around vortex during dissipation} 
\end{figure}

We also measured large increases in the Mach number near
the vortex as it reaches maximum $m=1$ amplitude and begins to decay. 
Figure \ref{machplot} plots the Mach number $M=|\bm{v} -
\bm{v}_\mathrm{vor}|/c_s$, where 
$\bm{v}_\mathrm{vor}$ corresponds to the bulk velocity of the vortex
around the disc. Values in Fig. \ref{machplot} have been averaged over
a region within $2H$ of the vortex centre. 
During the quasi-steady state the Mach number increase
steadily, and for all cases $M$ reaches a maximum at 
% with larger growth rate for higher $\tilde\beta$.
a time coincident with the start of vortex decay. {\bf check if
  statement is true}

Putting the above observations together, we suggest the decay
in the vortex amplitude is due to shock formation \emph{by the
  vortex}. When the vortex reaches large amplitude, it begins
to induce shocks in the surrounding fluid, as supported by the
increase in Mach number and the appearence of wakes with large surface
density gradients. The vortex may lose energy through shock
dissipation. In addition, a strong vortex (or shock formation) 
can smooth out the gap structure that originally gave rise to the
RWI, which would oppose vortex growth. We examine this below.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/mach}
  \caption{Running-time average {\bf ?} of the Mach number relative to
    the vortex, averaged over a region within $2H$ of the  
    vortex centre. This plot can be compared to the evolution of the
    vortex amplitude shown in Fig. \ref{lifetimeplot}.
    \label{machplot}}
\end{figure}
